Recursive Solutions to First-Order Model Counting

### 2023-01-23 doc/paper/paper.tex
Copy over some improvements from the thesis

% TODO: double check how the university names should be formatted and the emails
% TODO: read the formatting instructions and their LaTeX source
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore\\
$^2$University of Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s) \cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting \cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals \cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time \cite{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical \cite{DBLP:journals/ml/KimmigMG15} and statistical
relational \cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory \cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics \cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

% TODO: have ~ in front of every \cite
% TODO: fix doom emacs warnings
Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the domain(s)
\cite{jaeger2012liftability}. Over more than a decade, many classes of instances
were shown to be liftable
\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment \cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and \SRU{}
\cite{DBLP:conf/nips/KazemiKBP16}, \Uone{} \cite{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms \cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the empirical
front, there are several open-source implementations of exact WFOMC algorithms:
\textsc{ForcLift} \cite{DBLP:conf/ijcai/BroeckTMDR11}, probabilistic theorem
proving \cite{DBLP:journals/cacm/GogateD16}, and \textsc{L2C}
\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{ApproxWFOMC} \cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift} \cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving \cite{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is previous work
  \cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/ijcai/Broeck16} showing how FOMC
  can compute factorials in principle, but an algorithm that can lift such an
  instance in practice does not exist.} We claim that this shortcoming is due to
the inability of these algorithms to construct recursive solutions. The topic of
recursion in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

\subsection{Generalised Domain Recursion}\label{sec:dr}
% TODO: the thesis has some improvements to this section

@@

% Original domain recursion
Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} imposes stringent
preconditions on the input formula to ensure that the expanded version of the
formula (as in \cref{example:dr}) can be handled efficiently. For example, at
least one clause must contain a variable that occurs in all of its atoms. The
clauses in this expanded formula are then partitioned into three parts based on
whether the transformation introduced constants or constraints or both. The
preconditions ensure that these parts can be treated independently.

@@

\paragraph{Caching.}
\textsc{ForcLift} \cite{DBLP:conf/ijcai/BroeckTMDR11} uses a cache to check if
a formula is identical to one of the formulas that have already been fully
compiled. To facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

@@

When \textsc{ForcLift} \cite{DBLP:conf/ijcai/BroeckTMDR11} compiles a WFOMC
instance into a circuit, each node type encodes an arithmetic operation on its
inputs and parameters. These operations are then immediately performed while
traversing the circuit. With \textsc{Crane}, the interpretation of an FCG is a
collection of functions. Each function has (some) domain sizes as parameters and
may contain recursive calls to other functions, including itself. While there
may be any number of subsidiary functions, there is always one main function
that can be called with the sizes of the domains of the input formula as
arguments. Henceforth, this function is always called $f$, and it is defined by
the source node.

The interpretation of a node is decided by its type. Here we describe the
interpretations of new (or significantly changed) types and refer the reader to
previous work \cite{DBLP:conf/ijcai/BroeckTMDR11} for information on other
types. The interpretation of a $\CR$ or a $\GDR$ node is simply the
interpretation of its only direct successor. Obviously, $\star$ nodes have no
interpretation as incomplete FCGs are not meant to be interpreted. The
interpretation of a $\Reff$ node is a function call. The direct successor of the
$\Reff$ node (say, $v$) then must introduce a function. The parameters of this
function are the sizes of all domains used by nodes reachable from $v$.

@@

In this paper, we showed how a state-of-the-art (W)FOMC algorithm can be
empowered by generalising domain recursion and adding support for cycles in the
graph that encodes a solution. To construct such graphs, \textsc{Crane}
supplements \textsc{ForcLift} \cite{DBLP:conf/ijcai/BroeckTMDR11} with three
new compilation rules. Our experiments revealed a range of counting problems
that are liftable to \textsc{Crane} but not to any other (W)FOMC algorithm.
Although in this paper we focus on unweighted counting, \textsc{ForcLift}'s
support for weights trivially transfers to \textsc{Crane} as well. The common
thread across these newly liftable problems is (partial) injectivity. Thus, we
can formulate the following conjecture.

@@

However, the most important direction for future work is to fully automate this
new process. First, we need an algorithm that transforms FCGs into definitions
of functions. Formalising this process would also allow us to prove the
correctness of the new compilation rules in constructing FCGs that indeed
compute the right (weighted) model counts. Second, these definitions must be
simplified before they can be used, perhaps by a computer algebra system. Third,
we need a way to find the base cases for the recursive definitions provided by
\textsc{Crane}. Fourth, since the first solution found by \textsc{Crane} is not
always optimal in terms of its complexity, an automated way to determine the
asymptotic complexity of a solution would be helpful as well. Achieving these
goals would enable \textsc{Crane} to automatically constructing efficient ways
to compute various functions and sequences. In addition to the potential impact
on areas of artificial intelligence (AI) such as statistical relational AI
\cite{DBLP:series/synthesis/2016Raedt}, \textsc{Crane} could be beneficial to
research in combinatorics as well \cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

whose complexity is in \cref{tbl:results} (including those findable by
\textsc{ForcLift} \cite{DBLP:conf/ijcai/BroeckTMDR11}). Both before and after
algebraic simplification, excluding multiplication by one. All of these
solutions have been checked to be correct when combined with suitable base
cases.

>>>>

% TODO: double check how the university names should be formatted and the emails
% TODO: read the formatting instructions and their LaTeX source
% TODO: make sure there are no overflow errors
\author{%
Paulius Dilkas$^1$\and
Vaishak Belle$^2$\\
\affiliations
$^1$National University of Singapore, Singapore\\
$^2$University of Edinburgh, UK\\
\emails
paulius.dilkas@nus.edu.sg,
vbelle@ed.ac.uk
}

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic (FOL) given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances
become solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. Lifted
inference algorithms are typically used on probabilistic models such as
probabilistic programming languages
\cite{DBLP:journals/ml/RaedtK15,DBLP:journals/ijar/RiguzziBZCL17}, Markov logic
networks
\cite{DBLP:conf/ijcai/BroeckTMDR11,DBLP:journals/cacm/GogateD16,DBLP:journals/ml/RichardsonD06},
and other lifted graphical~\cite{DBLP:journals/ml/KimmigMG15} and statistical
relational~\cite{DBLP:series/synthesis/2016Raedt} models. Lifted inference
techniques for probabilistic databases, while developed somewhat independently,
have also been inspired by WFOMC
\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC tends to receive more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

% TODO: have ~ in front of every~\cite
% TODO: fix doom emacs warnings
Traditionally in computational complexity theory, a problem is \emph{tractable}
if it can be solved in time polynomial in the instance size. The equivalent
notion in (W)FOMC is liftability. A (W)FOMC instance is \emph{(domain-)liftable}
if it can be solved in time polynomial in the size(s) of the domain(s)
\cite{jaeger2012liftability}. Over more than a decade, many classes of instances
were shown to be liftable
\cite{DBLP:conf/kr/BremenK21,DBLP:conf/nips/KazemiKBP16,DBLP:conf/lics/KuusistoL18,DBLP:journals/jair/Kuzelka21}.
First, Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} showed that the class
of all sentences of FOL with up to two variables (denoted \FOtwo{}) is liftable.
Then Beame et al.~\shortcite{DBLP:conf/pods/BeameBGS15} proved that there exists
a sentence with three variables for which FOMC is $\#\P_{1}$-complete (i.e.,
\FOthree{} is not liftable). Since these two results came out, most of the
research on (W)FOMC focused on developing faster solutions for the \FOtwo{}
fragment~\cite{DBLP:conf/uai/BremenK21,DBLP:conf/aaai/MalhotraS22} and defining
new liftable fragments. These fragments include \SFO{} and \SRU{}
\cite{DBLP:conf/nips/KazemiKBP16}, \Uone{}~\cite{DBLP:conf/lics/KuusistoL18},
\FOtwo{} with tree axioms~\cite{DBLP:conf/kr/BremenK21}, and \Ctwo{} (i.e., the
two-variable fragment with counting quantifiers)
\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}. On the empirical
front, there are several open-source implementations of exact WFOMC algorithms:
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}, probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}, and \textsc{L2C}
\cite{DBLP:conf/kr/KazemiP16}. Approximate counting is supported by
\textsc{ApproxWFOMC}~\cite{DBLP:conf/ijcai/BremenK20} as well as
\textsc{ForcLift}~\cite{DBLP:conf/uai/BroeckCD12} and probabilistic theorem
proving~\cite{DBLP:journals/cacm/GogateD16}.

@@

However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is previous work
 ~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/ijcai/Broeck16} showing how FOMC
  can compute factorials in principle, but an algorithm that can lift such an
  instance in practice does not exist.} We claim that this shortcoming is due to
the inability of these algorithms to construct recursive solutions. The topic of
recursion in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

\subsection{Generalised Domain Recursion}\label{sec:dr}

@@

% Original domain recursion
Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11} imposes stringent conditions
on the input formula to ensure that the expanded version of the formula (as in
\cref{example:dr}) can be handled efficiently. For instance,
\cref{example:first} cannot be handled by \textsc{ForcLift} because there is no
root binding class, i.e., the two root variables belong to different equivalence
classes with respect to the binding relationship. The clauses in this expanded
formula are then partitioned into three parts based on whether the
transformation introduced constants or constraints or both. The aforementioned
conditions ensure that these parts can be treated independently.

@@

\paragraph{Caching.}
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} uses a cache to check if
a formula is identical to one of the formulas that have already been fully
compiled. To facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

@@

When \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} compiles a WFOMC
instance into a circuit, each node type encodes an arithmetic operation on its
inputs and parameters. These operations are then immediately performed while
traversing the circuit. With \textsc{Crane}, the interpretation of an FCG is a
collection of functions. Each function has (some) domain sizes as parameters and
may contain recursive calls to other functions, including itself. While there
may be any number of subsidiary functions, there is always one main function
that can be called with the sizes of the domains of the input formula as
arguments. Henceforth, this function is always called $f$, and it is defined by
the source node.

The interpretation of a node is decided by its type. Here we describe the
interpretations of new (or significantly changed) types and refer the reader to
previous work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information on other
types. The interpretation of a $\CR$ or a $\GDR$ node is simply the
interpretation of its only direct successor. Obviously, $\star$ nodes have no
interpretation as incomplete FCGs are not meant to be interpreted. The
interpretation of a $\Reff$ node is a function call. The direct successor of the
$\Reff$ node (say, $v$) then must introduce a function. The parameters of this
function are the sizes of all domains used by nodes reachable from $v$.

@@

In this paper, we showed how a state-of-the-art (W)FOMC algorithm can be
empowered by generalising domain recursion and adding support for cycles in the
graph that encodes a solution. To construct such graphs, \textsc{Crane}
supplements \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} with three
new compilation rules. Our experiments revealed a range of counting problems
that are liftable to \textsc{Crane} but not to any other (W)FOMC algorithm.
Although in this paper we focus on unweighted counting, \textsc{ForcLift}'s
support for weights trivially transfers to \textsc{Crane} as well. The common
thread across these newly liftable problems is (partial) injectivity. Thus, we
can formulate the following conjecture.

@@

However, the most important direction for future work is to fully automate this
new process. First, we need an algorithm that transforms FCGs into definitions
of functions. Formalising this process would also allow us to prove the
correctness of the new compilation rules in constructing FCGs that indeed
compute the right (weighted) model counts. Second, these definitions must be
simplified before they can be used, perhaps by a computer algebra system. Third,
we need a way to find the base cases for the recursive definitions provided by
\textsc{Crane}. Fourth, since the first solution found by \textsc{Crane} is not
always optimal in terms of its complexity, an automated way to determine the
asymptotic complexity of a solution would be helpful as well. Achieving these
goals would enable \textsc{Crane} to automatically constructing efficient ways
to compute various functions and sequences. In addition to the potential impact
on areas of artificial intelligence (AI) such as statistical relational AI
\cite{DBLP:series/synthesis/2016Raedt}, \textsc{Crane} could be beneficial to
research in combinatorics as well~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

@@

whose complexity is in \cref{tbl:results} (including those findable by
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}). Both before and after
algebraic simplification, excluding multiplication by one. All of these
solutions have been checked to be correct when combined with suitable base
cases.
